---
title: "Pima Diabetes Dataset Analysis"
author: "Krishna Kousalya Kesavan - Coding Project"
date: "`r Sys.Date()`"
format: 
  html:
    code-fold: true
    code-summary: "Show code"
    toc: true
    toc-depth: 2
    theme: cosmo
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 8,
  fig.height = 5
)
```

## Introduction

This analysis explores the Pima Indians Diabetes Dataset containing medical measurements for 768 female patients. The data was collected by the National Institute of Diabetes and Digestive and Kidney Diseases and includes diagnostic measurements to investigate diabetes risk factors. The Pima Indian population has historically shown high diabetes prevalence, making this dataset valuable for understanding disease predictors.
Data was originally collected in the 1990s from Pima Indian women aged 21+ living near Phoenix, Arizona.

**Research Questions:**

1. How do glucose levels differ between diabetes groups?
2. Which variables show the strongest differences?
3. Does age affect diabetes prevalence?
4. Which factor combinations appear most risky?

## Data Loading and Preparation

```{r load-packages}
library(dplyr)
library(ggplot2)
library(tidyr)
library(gridExtra)
```

```{r load-data}
diabetes_data <- read.csv("datafolder/Pima_diabetes.csv")

diabetes_data$Outcome <- factor(diabetes_data$Outcome, 
                                levels = c(0, 1), 
                                labels = c("No Diabetes", "Diabetes"))

# Data overview
summary_stats <- data.frame(
  Observations = nrow(diabetes_data),
  Variables = ncol(diabetes_data),
  Diabetes_Cases = sum(diabetes_data$Outcome == "Diabetes"),
  Prevalence_Percent = round(sum(diabetes_data$Outcome == "Diabetes") / nrow(diabetes_data) * 100, 1)
)

knitr::kable(summary_stats, caption = "Dataset Overview")
```

**Note on Data Quality:** Some variables contain physiologically implausible zero values (e.g., BloodPressure = 0), likely representing missing data. These are retained in the analysis but should be interpreted cautiously.

## Question 1: Glucose Levels and Diabetes

Blood glucose is the primary diagnostic criterion for diabetes (ADA threshold: ≥126 mg/dL fasting).

```{r glucose-analysis}
compare_groups <- function(data, variable, var_label) {
  summary_stats <- data %>%
    group_by(Outcome) %>%
    summarise(
      Mean = mean(get(variable), na.rm = TRUE),
      SD = sd(get(variable), na.rm = TRUE),
      Median = median(get(variable), na.rm = TRUE),
      Q25 = quantile(get(variable), 0.25, na.rm = TRUE),
      Q75 = quantile(get(variable), 0.75, na.rm = TRUE),
      .groups = "drop"
    )
  
  plot <- ggplot(data, aes(x = Outcome, y = get(variable), fill = Outcome)) +
    geom_violin(alpha = 0.6) +
    geom_boxplot(width = 0.2, alpha = 0.8) +
    labs(
      title = paste(var_label, "by Diabetes Status"),
      y = var_label,
      x = "Diabetes Status"
    ) +
    theme_minimal() +
    scale_fill_manual(values = c("#4ECDC4", "#FF6B6B")) +
    theme(legend.position = "none")
  
  return(list(stats = summary_stats, plot = plot))
}

glucose_results <- compare_groups(diabetes_data, "Glucose", "Glucose Level (mg/dL)")

knitr::kable(glucose_results$stats, 
             caption = "Glucose Level Summary Statistics",
             digits = 1)

glucose_results$plot
```

**Findings:** While mean glucose differs substantially between groups (~31 mg/dL), considerable overlap exists. Notably, 25% of non-diabetics have glucose levels exceeding 113 mg/dL, while 25% of diabetics have levels below 111 mg/dL, indicating significant distributional overlap. Glucose alone provides incomplete classification—a person with 150 mg/dL could belong to either group. This suggests diabetes diagnosis requires consideration of multiple factors beyond a single glucose measurement.

## Question 2: Comparing Multiple Variables

To objectively compare different variables measured on different scales, we use Cohen's d effect size (standardized difference between group means).

**Effect size interpretation:** Small (0.2), Medium (0.5), Large (0.8+)

```{r effect-sizes}
# Cohen's d calculation
cohens_d <- function(group1, group2) {
  mean_diff <- mean(group1, na.rm = TRUE) - mean(group2, na.rm = TRUE)
  pooled_sd <- sqrt((var(group1, na.rm = TRUE) + var(group2, na.rm = TRUE)) / 2)
  return(mean_diff / pooled_sd)
}

variables_to_compare <- c("Glucose", "BMI", "Age", "Pregnancies", 
                         "BloodPressure", "Insulin")

effect_sizes <- sapply(variables_to_compare, function(var) {
  no_diabetes <- diabetes_data[diabetes_data$Outcome == "No Diabetes", var]
  diabetes <- diabetes_data[diabetes_data$Outcome == "Diabetes", var]
  abs(cohens_d(no_diabetes, diabetes))
})

effect_size_results <- data.frame(
  Variable = variables_to_compare,
  Effect_Size = effect_sizes
) %>%
  arrange(desc(Effect_Size)) %>%
  mutate(Interpretation = case_when(
    Effect_Size >= 0.8 ~ "Large",
    Effect_Size >= 0.5 ~ "Medium",
    Effect_Size >= 0.2 ~ "Small",
    TRUE ~ "Negligible"
  ))

knitr::kable(effect_size_results, 
             caption = "Variables Ranked by Effect Size",
             digits = 3)

# Visualize with interpretation markers
ggplot(effect_size_results, aes(x = reorder(Variable, Effect_Size), 
                                y = Effect_Size, fill = Interpretation)) +
  geom_col(alpha = 0.8) +
  coord_flip() +
  geom_hline(yintercept = 0.8, linetype = "dashed", color = "darkgreen", alpha = 0.7) +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "orange", alpha = 0.7) +
  geom_hline(yintercept = 0.2, linetype = "dashed", color = "red", alpha = 0.7) +
  annotate("text", x = 6.5, y = 0.85, label = "Large (0.8)", size = 3, hjust = 0) +
  annotate("text", x = 6.5, y = 0.55, label = "Medium (0.5)", size = 3, hjust = 0) +
  annotate("text", x = 6.5, y = 0.25, label = "Small (0.2)", size = 3, hjust = 0) +
  labs(
    title = "Strength of Association with Diabetes",
    x = "Variable",
    y = "Effect Size (Cohen's d)"
  ) +
  theme_minimal() +
  scale_fill_manual(values = c("Large" = "#2E7D32", "Medium" = "#F57C00", "Small" = "#C62828")) +
  theme(legend.position = "bottom")
```

**Findings:** Glucose shows the strongest association (large effect), followed by BMI (medium effect). Age and Pregnancies show smaller effects, while BloodPressure and Insulin show weak associations. The hierarchy suggests glucose and BMI are the most important discriminators, but even the strongest predictor (glucose) has limitations in classification due to group overlap.

## Question 3: Age and Diabetes Prevalence

```{r age-prevalence}
age_prev <- diabetes_data %>%
  mutate(AgeGroup = cut(Age, 
                       breaks = c(0, 30, 40, 50, 100),
                       labels = c("21-30", "31-40", "41-50", "50+"))) %>%
  group_by(AgeGroup) %>%
  summarise(
    Total = n(),
    Diabetes_Cases = sum(Outcome == "Diabetes"),
    Prevalence = Diabetes_Cases / Total * 100,
    .groups = "drop"
  )

knitr::kable(age_prev, 
             caption = "Diabetes Prevalence by Age Group",
             digits = 1)

ggplot(age_prev, aes(x = AgeGroup, y = Prevalence, fill = AgeGroup)) +
  geom_col(alpha = 0.8) +
  geom_text(aes(label = paste0(round(Prevalence, 1), "%")), 
            vjust = -0.5, size = 4) +
  labs(
    title = "Diabetes Prevalence by Age Group",
    x = "Age Group",
    y = "Prevalence (%)"
  ) +
  theme_minimal() +
  scale_fill_brewer(palette = "Set2") +
  guides(fill = "none")
```

**Findings:** Prevalence increases dramatically with age, from 21.6% in the youngest group to 56.6% in ages 41-50. This extremely high prevalence (35% overall, >50% in older groups) is characteristic of the Pima population, which has among the world's highest diabetes rates due to genetic susceptibility and rapid lifestyle changes. These rates are not representative of the general population.

## Question 4: High-Risk Factor Combinations

We examine two-dimensional relationships to identify high-risk profiles. Smoothed trend lines show how diabetes probability varies across the risk space.

```{r risk-combinations, fig.width=10, fig.height=8}

create_risk_profile <- function(data, var1, var2, label1, label2) {
  ggplot(data, aes(x = get(var1), y = get(var2), color = Outcome)) +
    geom_point(alpha = 0.5, size = 2) +
    geom_smooth(method = "loess", se = FALSE, linewidth = 1.2) +
    labs(
      title = paste(label1, "vs", label2),
      x = label1,
      y = label2,
      color = "Outcome"
    ) +
    theme_minimal() +
    scale_color_manual(values = c("#4ECDC4", "#FF6B6B"))
}

p1 <- create_risk_profile(diabetes_data, "Glucose", "BMI", 
                         "Glucose (mg/dL)", "BMI")
p2 <- create_risk_profile(diabetes_data, "Age", "BMI", 
                         "Age (years)", "BMI")
p3 <- create_risk_profile(diabetes_data, "Glucose", "Age", 
                         "Glucose (mg/dL)", "Age (years)")
p4 <- create_risk_profile(diabetes_data, "Pregnancies", "Age", 
                         "Number of Pregnancies", "Age (years)")

grid.arrange(p1, p2, p3, p4, ncol = 2)
```

**Findings:** The smoothed curves (LOESS regression) show non-linear relationships between variables and help identify high-risk regions. The glucose-BMI plot shows strongest separation: individuals with both high glucose (>140 mg/dL) and high BMI (>35) have highest diabetes probability. The age-BMI combination also shows risk amplification at higher values. Pregnancy count shows weaker discriminative ability when combined with age, suggesting it adds limited information beyond age alone.

## Summary and Conclusions

**Key Findings:**

1. **Glucose** is the strongest single predictor (effect size: 1.07) but shows substantial group overlap, indicating diabetes classification requires multiple factors
2. **BMI** is an important modifiable risk factor (effect size: 0.65) 
3. **Age-related risk** is pronounced, though the extremely high prevalence (35% overall, >50% in older groups) reflects this specific high-risk population
4. **Combined risk factors** create multiplicative effects—high glucose + high BMI + advanced age represents the highest-risk profile
5. **Single-variable limitations**: Even the strongest predictor cannot perfectly classify individuals, supporting the need for multivariable approaches (e.g., logistic regression models)

**Dataset Representativeness:** This dataset captures a population with genetic susceptibility and historically high diabetes rates. The 35% prevalence far exceeds the ~10% US national average, limiting generalizability. However, it remains valuable for studying risk factor relationships in high-risk populations.

**Data Quality Considerations:** Physiologically implausible zero values in BloodPressure and Insulin suggest missing data coded as zeros. Future analyses should properly handle these missing values.

---

## Appendix: Additional Variables

### BMI Analysis

```{r bmi-analysis}
bmi_results <- compare_groups(diabetes_data, "BMI", "Body Mass Index")

knitr::kable(bmi_results$stats, 
             caption = "BMI Summary Statistics",
             digits = 1)

bmi_results$plot
```

**Findings:** The diabetes group shows mean BMI ~4.8 points higher (obese category) versus non-diabetes group (overweight category). This supports weight management as a prevention strategy, though overlap indicates BMI alone is insufficient for risk stratification.

### Pregnancies Analysis

```{r pregnancy-analysis}
pregnancy_results <- compare_groups(diabetes_data, "Pregnancies", 
                                   "Number of Pregnancies")

knitr::kable(pregnancy_results$stats, 
             caption = "Pregnancy Summary Statistics",
             digits = 1)

pregnancy_results$plot
```

**Findings:** Women with diabetes average 1.6 more pregnancies. The notably high pregnancy counts (mean 3.3-4.9) reflect both the population's demographics and data collection from the 1990s. This association may be confounded by age and historical GDM, requiring cautious interpretation.

### Blood Pressure Analysis

```{r bp-analysis}
bp_results <- compare_groups(diabetes_data, "BloodPressure", 
                            "Blood Pressure (mm Hg)")

knitr::kable(bp_results$stats, 
             caption = "Blood Pressure Summary Statistics",
             digits = 1)

bp_results$plot
```

**Findings:** Modest difference between groups (2.6 mm Hg), with weak effect size (0.13). **Data quality concern**: Zero values likely represent missing data rather than true measurements. Blood pressure shows limited discriminative power for diabetes status in this dataset.

### Insulin Analysis

```{r insulin-analysis}
# Exclude zero values (missing data)
insulin_nonzero <- diabetes_data %>% filter(Insulin > 0)

insulin_results <- compare_groups(insulin_nonzero, "Insulin", 
                                 "Insulin Level (μU/mL)")

knitr::kable(insulin_results$stats, 
             caption = "Insulin Summary Statistics (non-zero values only)",
             digits = 1)

insulin_results$plot
```

**Findings:** After excluding 374 zero values (48.7% missing), insulin shows high variability in both groups. The weak effect size (0.26) and data quality issues limit its utility. Insulin's dynamic nature (varies with meals, disease stage) makes it challenging to interpret from single measurements.

